Supplementary material

### 2022-04-10 analysis/paper/paper.Rmd
Adding the Pauler data to all the other pre-reported shoreline dates and elevations

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b, see also @jorgensen2020; @tallavaara2020]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].

@@

As the excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure, the area immediately surrounding the sites has sometimes been severely impacted by modern disturbances. In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension [e.g. @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021]. In addition to code and original spatial data being available in the digital research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.

>>>>

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @jaksland2014; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b; see also @jorgensen2020; @tallavaara2020]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].

@@

The excavation of archaeological sites typically follow from residential and commercial development, as well as the expansion of infrastructure. As the data collection for the DTM was started by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has therefore sometimes been severely impacted by disturbances after the excavation of the site. In addition to employing 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension [e.g. @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021]. In addition to code and original spatial data being available in the digital research compendium for this paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.

### 2022-04-10 analysis/supplementary-materials/supplementary.Rmd
Adding the Pauler data to all the other pre-reported shoreline dates and elevations

@@

>>>>

Typological indicators in the lithic inventory from Brunstad 24 matches the ^14^C-dates [@danielsen2018]. The footprint of a large building structure covering a large part of the site from the south required editing of the DTM. Apart from that the landscape appears relatively undisturbed. 8 fragments from a pot could potentially correspond with the ^14^C-dates to the Late Bronze Age and Early Iron Age, but given their uncetaint nature and the limited number of sherds this was disregarded here.

@@

As with Brunstad 24 (above), the artefact inventory from Brunstad 25 matches the radiocarbon dates [@reitan2018], and the building structure covering the southern part of the site had impacted the DTM and had to be removed.
